---
title: "Untitled"
format: html
editor: visual
---

```{r}
library(ggplot2)
library(brms)
library(ggeffects)
library(tidyverse)
library(readxl)
library(here)
library(dplyr)
#install.packages("ggtext")
library(ggtext)
krat <- read.csv(here("final_krat_data_2026.csv"))
#print(krat)
```

## Measurements:

-   Body length

-   Tail length

-   Hind limb length

-   Head length

-   Foot length

-   Ear length

-   Foot Width

-   Wrist Width

-   Finger length

-   Hand Width

-   Toe length

## Steps to analyze

1.  Ratio of head length to body length as a proxy for size

    ```{r}
    #Creating a new column to use as a proxy for the size of the individual by dividing head_length by body_length
    df <- krat %>%
        mutate(relative_size = Body_Length / Head_Length)
    ```

<!-- -->

2.  plot each measurement vs. size proxy

<!-- -->

3.  Analyze with bayesian bernoulli model.

<!-- -->

4.  Tested the Predictive Power

## Body length vs Relative size

```{r}
#Plot the body length vs Relative size
df %>%
  ggplot(aes(y= Body_Length, x = relative_size, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")

#Running bernouilli model to predict species based on measurements
m.bern.body <- brm(Species ~ 1 + Body_Length + relative_size, 
                        data = df, 
                        family = bernoulli(), 
                        chains = 4, iter = 2000,
                        file = "output/m.bern.body") 

#summary model
summary(m.bern.body)
plot(m.bern.body)

#Predict the response based on the model and plot it
pred.m.body <- predict_response(m.bern.body)
plot(pred.m.body, show_data = TRUE)

#Check the probability of each species
pp_check(m.bern.body, type="scatter_avg")

#Use function to check if these were right
fitted(m.bern.body)
```

### Interpretation:

confidence interval above zero and the estimate is 0.40. Looking at the predicted probabilities of species vs body length there are extreme values where we can be sure of species, but the middle values have overlap which can be uncertain between the two species.

## Tail length vs Relative Size

```{r}
#Plot the Tail length vs Relative size
df %>%
  ggplot(aes(y= Tail_Length, x = relative_size, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")

#Running bernouilli model to predict species based on measurements
m.bern.tail <- brm(Species ~ 1 + Tail_Length + relative_size, 
                        data = df, 
                        family = bernoulli(), 
                        chains = 4, iter = 2000,
                        file = "output/m.bern.tail") 

#summary model
summary(m.bern.tail)
plot(m.bern.tail)

#Predict the response based on the model and plot it
preds.m.tail <- predict_response(m.bern.tail)
plot(preds.m.tail, show_data = TRUE)

#Check the probability of each species
pp_check(m.bern.tail, type="scatter_avg")

#Use function to check if these were right
fitted(m.bern.tail)
```

### Interpretation:

confidence interval above zero and the estimate is 0.16. Looking at the predicted probabilities of species vs tail length there are extreme values where we can be sure of species, but the middle values have overlap which can be uncertain between the two species.

## Head Length vs Relative size

```{r}
#Plot the head length vs Relative size
df %>%
  ggplot(aes(y= Head_Length, x = relative_size, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")

#Running bernouilli model to predict species based on measurements
m.bern.head2 <- brm(Species ~ 1 + Head_Length + relative_size, 
                        data = df, 
                        family = bernoulli(), 
                        chains = 4, iter = 2000,
                        file = "output/m.bern.head2") 

#summary model
summary(m.bern.head2)
plot(m.bern.head2)

#Predict the response based on the model and plot it
preds.m.head2 <- predict_response(m.bern.head2)
plot(preds.m.head2, show_data = TRUE)

#Check the probability of each species
pp_check(m.bern.head2, type="scatter_avg")

#Use function to check if these were right
fitted(m.bern.head2)
```

### Interpretation:

The confidence interval does not include 0 and it has an estimate of 0.98. Looking in the predicted probabiltiy of species vs head length at the extreme values we can be sure that it is one species or the other, however in the overlap it can be harder to tell the species.

## Hindlimb length vs Relative size

```{r}
#Plot the hindlimb vs Relative size
df %>%
  ggplot(aes(y= Hind_Limb_Length, x = relative_size, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")

#Running bernouilli model to predict species based on measurements
m.bern.hindlimb <- brm(Species ~ 1 + Hind_Limb_Length + relative_size, 
                        data = df, 
                        family = bernoulli(), 
                        chains = 4, iter = 2000,
                        file = "output/m.bern.hindlimb") 

#summary model
summary(m.bern.hindlimb)
plot(m.bern.hindlimb)

#Predict the response based on the model and plot it
preds.m.hindlimb <- predict_response(m.bern.hindlimb)
plot(preds.m.hindlimb, show_data = TRUE)

#Check the probability of each species
pp_check(m.bern.hindlimb, type="scatter_avg")

#Use function to check if these were right
fitted(m.bern.hindlimb)
```

### Interpretation:

This seems like it could be a good indicator with a confidence interval greater than 0 and an estimate of 1.13 which is pretty big. Looking at the predicted probabilities of species of the hind limb length we can be sure about the species for extreme values for each species, however there is a large overlap in the values.

## Hindlimb length vs hindlimb/bodysize

```{r}
df2 <- krat %>%
    mutate(relative_hindlimb = Body_Length / Hind_Limb_Length)

#Plot the hindlimb vs Relative size
df2 %>%
  ggplot(aes(y= Hind_Limb_Length, x = relative_hindlimb, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")

#Running bernouilli model to predict species based on measurements
m.bern.hindlimb3 <- brm(Species ~ 1 + Hind_Limb_Length + relative_hindlimb, 
                        data = df2, 
                        family = bernoulli(), 
                        chains = 4, iter = 2000,
                        file = "output/m.bern.hindlimb2") 

#summary model
summary(m.bern.hindlimb3)
plot(m.bern.hindlimb3)

#Predict the response based on the model and plot it
preds.m.hindlimb3 <- predict_response(m.bern.hindlimb3)
plot(preds.m.hindlimb3, show_data = TRUE)

#Check the probability of each species
pp_check(m.bern.hindlimb3, type="scatter_avg")

#Use function to check if these were right
fitted(m.bern.hindlimb3)
```

### Interpretation:

Confidence interval does not include 0, and the estimate was 1.08. There are extreme values where we are sure of species but still plenty of overlap in the middle values where we can not be sure.

## Foot length vs Relative Size

```{r}
#Plot the foot length vs Relative size
df %>%
  ggplot(aes(y= Foot_Length, x = relative_size, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")

#Running bernouilli model to predict species based on measurements
m.bern.footL <- brm(Species ~ 1 + Foot_Length + relative_size, 
                        data = df, 
                        family = bernoulli(), 
                        chains = 4, iter = 2000,
                        file = "output/m.bern.footL") 

#summary model
summary(m.bern.footL)
plot(m.bern.footL)

#Predict the response based on the model and plot it
preds.m.footL <- predict_response(m.bern.footL)
plot(preds.m.footL, show_data = TRUE)

#Check the probability of each species
pp_check(m.bern.footL, type="scatter_avg")

#Use function to check if these were right
fitted(m.bern.footL)
```

### Interpretation:

Foot length estimates confidence interval does not include 0 and the value is 1.27. The predicted probability of species graph shows that there are values of foot length where we can be 100% certain of species and some where there is overlap.

## Ear Length vs Relative size

```{r}
#Plot the Ear length vs Relative size
df %>%
  ggplot(aes(y= Ear_Length_Notch, x = relative_size, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")+ylim(0,20)


#Running bernouilli model to predict species based on measurements
m.bern.ear <- brm(Species ~ 1 + Ear_Length_Notch + relative_size, 
                        data = df, 
                        family = bernoulli(), 
                        chains = 4, iter = 2000,
                        file = "output/m.bern.ear") 

#summary model
summary(m.bern.ear)
plot(m.bern.ear)

#Predict the response based on the model and plot it
preds.m.ear <- predict_response(m.bern.ear)
plot(preds.m.ear, show_data = TRUE)

#Check the probability of each species
pp_check(m.bern.ear, type="scatter_avg")

#Use function to check if these were right
fitted(m.bern.ear)
```

### Interpretation:

Confidence interval includes 0 therefore we cannot be sure that the effect is different than 0. The predicted probabilities of species when looking at ear length is below 50% certainty which is not very good.

## Ear length vs Head length

```{r}
#Plot the Ear length vs the head length
df %>%
  ggplot(aes(y= Ear_Length_Notch, x = Head_Length, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")+ylim(10,20)

#Running bernouilli model to predict species based on measurements
m.bern.ear2 <- brm(Species ~ 1 + Ear_Length_Notch + Head_Length, 
                        data = df, 
                        family = bernoulli(), 
                        chains = 4, iter = 2000,
                        file = "output/m.bern.ear2") 

#summary model
summary(m.bern.ear2)
plot(m.bern.ear2)

#Predict the response based on the model and plot it
preds.m.ear2 <- predict_response(m.bern.ear2)
plot(preds.m.ear2, show_data = TRUE)

#Check the probability of each species
pp_check(m.bern.ear2, type="scatter_avg")

#Use function to check if these were right
fitted(m.bern.ear2)
```

### Interpretation:

Ear length vs head length bernouilli model outputs confidence interval for ear length include 0 which mean the probability could mean nothing.

## Foot Width vs Relative size

```{r}
#Plot the Foot Width vs Relative size
df %>%
  ggplot(aes(y= Foot_Width, x = relative_size, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")

#Running bernouilli model to predict species based on measurements
m.bern.footW <- brm(Species ~ 1 + Foot_Width + relative_size, 
                        data = df, 
                        family = bernoulli(), 
                        chains = 4, iter = 2000,
                        file = "output/m.bern.footW") 

#summary model
summary(m.bern.footW)
plot(m.bern.footW)

#Predict the response based on the model and plot it
preds.m.footW <- predict_response(m.bern.footW)
plot(preds.m.footW, show_data = TRUE)

#Check the probability of each species
pp_check(m.bern.footW, type="scatter_avg")

#Use function to check if these were right
fitted(m.bern.footW)
```

### Interpretation:

Could be a good indicator bc it does not include 0 in CI and the prediction graph shows a high and low probability. Value is 5.94 which is really big slope estimate.

## Wrist Width vs. Relative size

```{r}
#Plot the Wrist Width vs Relative size
df %>%
  ggplot(aes(y= Wrist_Width, x = relative_size, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")+ theme_bw()+
  labs(
    x="Relative size (Head length / Body length)",
    y="Wrist Width",
    title="Wrist Wisth over Relative Size of <br>*Dipodomys heermanni* and *Dipodomys venustus*"
  )+ theme(
    legend.text=element_text(face="italic"),
    legend.position=c(0.85,0.8),
     plot.title=element_markdown(hjust=0.5))+ 
  scale_color_manual(values=c("darkturquoise","palevioletred"))


#setwd("C:/Users/macke/Desktop/BIOE 176 - Data Science/RC-2026-Kangaroo-Rat-Project")


#Running bernouilli model to predict species based on measurements
m.bern.wrist <- brm(Species ~ 1 + Wrist_Width + relative_size, 
                        data = df, 
                        family = bernoulli(), 
                        chains = 4, iter = 2000,
                        file = "C:/Users/macke/Desktop/BIOE 176 - Data Science/RC-2026-Kangaroo-Rat-Project/output/m.bern.wrist") 

#summary model
summary(m.bern.wrist)
plot(m.bern.wrist)

#Predict the response based on the model and plot it
preds.m.wrist <- predict_response(m.bern.wrist)
plot(preds.m.wrist, show_data = TRUE)

#preds <- ggpredict(m.bern.wrist, terms = c("relative_size", "Wrist_Width"))

#ggplot(preds, aes(x, predicted, color = Species)) +
  #geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.25) +
  #geom_line(size = 1.2) +
  #geom_point(data = df, aes(x = relative_size, y = as.numeric(Species)), 
             #alpha = 0.7, linewidth = 2.5, position = position_jitter(width = 0.02)) +
  #scale_color_manual(values = c("darkturquoise", "palevioletred")) +
  #labs(x = "Relative size", y = "Probability", color = "Species") +
  #theme_minimal(base_size = 14)

#plot(conditional_effects(m.bern.wrist), 
     #rug = TRUE,
     #points = list(alpha = 0.6, size = 2.5)) +
  #scale_color_manual(values = c("darkturquoise", "palevioletred")) +
  #theme_minimal(base_size = 14) +
  #theme(legend.position = c(0.85, 0.8))

#plot(preds.m.wrist, show_data = TRUE) +
  #theme_minimal(base_size = 14) +
  #scale_color_manual(values = c("darkturquoise", "palevioletred")) +
  #theme(
    #legend.position = c(0.85, 0.8),
    #legend.background = element_rect(fill = "white", color = NA),
    #panel.grid.minor = element_blank()
  #) +
  #labs(
    #x = "Relative size (Head length / Body length)",
    #y = "Probability of Species",
    #title = "Wrist Width vs Relative Size Predictions"
  #)+geom_jitter()

#plot(conditional_effects(m.bern.wrist), 
     #rug = TRUE,
     #points = list(alpha = 0.7, size = 2.5)) +
  #scale_color_manual(values = c("darkturquoise", "palevioletred")) +
  #theme_minimal(base_size = 14) +
  #theme(legend.position = c(0.85, 0.8))

#plot(preds.m.wrist, show_data = TRUE) +
  #scale_color_manual(values = c("darkturquoise", "palevioletred")) +
  #theme_minimal(base_size = 14) +
  #theme(legend.position = c(0.85, 0.8))

#Check the probability of each species
pp_check(m.bern.wrist, type="scatter_avg")

#Use function to check if these were right
fitted(m.bern.wrist)
```

### Interpretation:

Includes 0 in the CI for the slope, therefore not good

## Finger length vs. Relative size

```{r}
#Plot the Finger length vs Relative size
df %>%
  ggplot(aes(y= Finger_Length, x = relative_size, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")

#Running bernouilli model to predict species based on measurements
m.bern.finger <- brm(Species ~ 1 + Finger_Length + relative_size, 
                        data = df, 
                        family = bernoulli(), 
                        chains = 4, iter = 2000,
                        file = "C:/Users/macke/Desktop/BIOE 176 - Data Science/RC-2026-Kangaroo-Rat-Project/output/m.bern.finger") 

#summary model
summary(m.bern.finger)
plot(m.bern.finger)

#plot(conditional_effects(m.bern.finger), 
    # rug = TRUE,
     #points = list(alpha = 0.6, size = 2.5)) +
  #scale_color_manual(values = c("darkturquoise", "palevioletred")) +
  #theme_minimal(base_size = 14) +
  #theme(legend.position = c(0.85, 0.8))

#plot(preds.m.finger, show_data = TRUE) +
  #theme_minimal(base_size = 14) +
  #scale_color_manual(values = c("darkturquoise", "palevioletred")) +
  #theme(
    #legend.position = c(0.85, 0.8),
    #legend.background = element_rect(fill = "white", color = NA),
    #panel.grid.minor = element_blank()
  #) +
  #labs(
    #x = "Relative size (Head length / Body length)",
    #y = "Probability of Species",
    #title = "Wrist Width vs Relative Size Predictions"
  #)+geom_jitter()

#Predict the response based on the model and plot it
preds.m.finger <- predict_response(m.bern.finger)
plot(preds.m.finger, show_data = TRUE)

#Check the probability of each species
pp_check(m.bern.finger, type="scatter_avg")

#Use function to check if these were right
fitted(m.bern.finger)
```

### Interpretation:

Finger length does not include 0 in CI and estimate is 3.64 which means its good??? has very distinct 100% certainty of species for certain values with only 2 that have a probability of being either.

## Hand width vs Relative size

```{r}
#Plot the Hand width vs Relative size
df %>%
  ggplot(aes(y= Hand_Width, x = relative_size, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")

#Running bernouilli model to predict species based on measurements
m.bern.handW <- brm(Species ~ 1 + Hand_Width + relative_size, 
                        data = df, 
                        family = bernoulli(), 
                        chains = 4, iter = 2000,
                        file = "output/m.bern.handW") 

#summary model
summary(m.bern.handW)
plot(m.bern.handW)

#Predict the response based on the model and plot it
preds.m.handW <- predict_response(m.bern.handW)
plot(preds.m.handW, show_data = TRUE)

#Check the probability of each species
pp_check(m.bern.handW, type="scatter_avg")

#Use function to check if these were right
fitted(m.bern.handW)
```

### Interpretation:

Confidence interval for hand width hovers around 0 and includes it meaning that there is no difference and this measurement has zero difference between each species.

## Toe Length vs. Relative size

```{r}
#Plot the Toe Length vs Relative size
df %>%
  ggplot(aes(y= Toe_Length, x = relative_size, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")

#Running bernouilli model to predict species based on measurements
m.bern.toe <- brm(Species ~ 1 + Toe_Length + relative_size, 
                        data = df, 
                        family = bernoulli(), 
                        chains = 4, iter = 2000,
                        file = "output/m.bern.toe") 

#summary model
summary(m.bern.toe)
plot(m.bern.toe)

#Predict the response based on the model and plot it
preds.m.toe <- predict_response(m.bern.toe)
plot(preds.m.toe, show_data = TRUE)

#Check the probability of each species
pp_check(m.bern.toe, type="scatter_avg")

#Use function to check if these were right
fitted(m.bern.toe)
```

### Interpretation:

Confidence interval does not include 0 meaning it can mean something. There is an upward slope in toe length and certainty of species, however we can be 100% certain about one species for some of the values but only 75% certain for the values in the other species.
