---
title: "Presentation visualizations"
format: html
editor: visual
---

## Measurements to present:

1\) Finger length

2\) Hand Width

## Goals:

1\) Use generated models to analyze the data

2\) Create graphs

3\) Format the graphs for presentations

4\) Make sure our interpretation is sound and concise

## Set Up:

```{r}
#load the packages necessary
library(ggplot2)
library(brms)
library(ggeffects)
library(tidyverse)
library(readxl)
library(here)
library(ggtext)
#Load the data
krat <- read.csv(here("final_krat_data_2026.csv"))
#print(krat)
```

## Good measurement: Finger length

```{r}
install.packages("ggtext")
library(ggtext)
#Plot the Finger length vs Relative size
df %>%
  ggplot(aes(y= Finger_Length, x = relative_size, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")+ theme_bw()+
  theme(
    legend.text=element_text(face="italic"),
    legend.position=c(0.85,0.2),
    plot.title=element_markdown(hjust=0.5)
        )+
  scale_color_manual(values=c("darkturquoise","palevioletred"))+
  labs(
    x="Relative size of head (Body length / Head length)",
    y="Finger length",
    title="Finger Length over Relative Head Size of 
    <br>*Dipodomys heermanni* and *Dipodomys venustus*"
  )

#Running bernouilli model to predict species based on measurements
m.bern.finger <- brm(Species ~ 1 + Finger_Length + relative_size, 
                        data = df, 
                        family = bernoulli(), 
                        chains = 4, iter = 2000,
                        file = "output/m.bern.finger") 
#summary model
summary(m.bern.finger)
plot(m.bern.finger)

#Predict the response based on the model and plot it
preds.m.finger <- predict_response(m.bern.finger)
plot(preds.m.finger, show_data = TRUE)

#Check the probability of each species
pp_check(m.bern.finger, type="scatter_avg")

#Use function to check if these were right
fitted(m.bern.finger)
```

\
xxx

```{r}
#Trying the bernouilli model without the proxy for size to see if its better
#run bernouilli model with only the finger length
m.bern.finger2 <- brm(Species ~ 1 + Finger_Length, 
                        data = df, 
                        family = bernoulli(), 
                        chains = 4, iter = 2000,
                        file = "output/m.bern.finger2") 
#summary model
summary(m.bern.finger2)
plot(m.bern.finger2)

#Predict the response based on the model and plot it
preds.m.finger2 <- predict_response(m.bern.finger2)
plot(preds.m.finger2, show_data = TRUE)

#Check the probability of each species
pp_check(m.bern.finger2, type="scatter_avg")

#Use function to check if these were right
fitted(m.bern.finger2)
```

### Interpretation:

Finger length does not include 0 in CI and estimate is 3.64 which means its good??? has very distinct 100% certainty of species for certain values with only 2 that have a probability of being either.

### How to make the visualizations better:

1\) stagger/jitter the values because they are stacked on like 5 values

2\) New colors + theme, etc

3\) Mutate to create table with probability of either species compared to actual species– make a histogram of what is correct??? to test how good our model is??

## Bad Measurement: Hand Width

```{r}
#Plot the Hand width vs Relative size
df %>%
  ggplot(aes(y= Hand_Width, x = relative_size, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")

#Running bernouilli model to predict species based on measurements
m.bern.handW <- brm(Species ~ 1 + Hand_Width + relative_size, 
                        data = df, 
                        family = bernoulli(), 
                        chains = 4, iter = 2000,
                        file = "output/m.bern.handW") 

#summary model
summary(m.bern.handW)
plot(m.bern.handW)

#Predict the response based on the model and plot it
preds.m.handW <- predict_response(m.bern.handW)
plot(preds.m.handW, show_data = TRUE)

#Check the probability of each species
pp_check(m.bern.handW, type="scatter_avg")

#Use function to check if these were right
fitted(m.bern.handW)
```

### Interpretation:

Confidence interval for hand width hovers around 0 and includes it meaning that there is no difference and this measurement has zero difference between each species.

### How to make the visualizations better:

1\) stagger the values because they are stacked on like 3 values

2\) New colors + theme, etc
