---
title: "KRAT DATA WRANGLING!!!!"
format: html
editor: visual
---

## RAW DATA!!!

```{r}
#installed readxl packages with #install.packages("read.xl"), & installed #rtools4.3 from online
#using the package
library(readxl)
#reading my tables and assigning them to a dataframe
table1 <- read_excel("postnovRAWkratdata.xlsx")
library(readxl)
table2 <- read_excel("PrenovRAWkratdata.xlsx")


```

Goal: these tables need to be put together, combined and tidy-ed.

They have different columns, table2 missing 2 diff headwidth (bullae (b or 2nd hw) and then maxillary (sometimes written as m or pm or 1st hw). Table2 also missing a tissue sample column. These values are written in the notes column in Table 2, but should be tidy in Table 1.

The animal weight in table 2 if it has a - after it means it needs to have about 16-20 grams subtracted from it. Also the math needs to be done in some columns to calculate animal weight.

animal weight = animal + bag weight - bag weight

sometimes the bag weight is missing so we estimate it as somewhere btwn 16-20 grams.

If there is an NA its because the data wasn't taken.

In certain columns NA is really just a default for something:

Column: repro stage NA = non-repro

Column: life stage NA = adult

Column: recap NA = not recap AKA a new individual

Column: Tissue sample NA = no tissue sample was taken

column: Location = WILL NEED TO BE FIGURED OUT USING GPS COORDS

The teeth column can be deleted from both data sets, it was not used.

The genus is correct for all, but the species is wrong. The species are based off of their location. We will have to use GPS coordinates to figure out species. With every krat from Fort Ord Natural Reserve, Marina, CA being a Dipodomys Heermanni (Genus: Di, Species: He) and all krats from Hastings Natural Reserve, Carmel Valley, CA being a Dipodomys Venustus (Genus: Di, Species: Ve).

There are some recaps, which we have tagged and measured before and then caught again which won't have any data except location associated with them in that recap observation. In our data analysis we should select for the first capture that includes the measurements and will have to make R ok with us having a bunch of NAs in our data set.

# Clean data

About a month ago I went through and tidied the data via excel spreadsheets as well as combining the two tables and addressing most of the things above.

It is however, still not correctly sorted on species and still contains the teeth column.

It may contain some of my human errors during the excel work.

It also does not have the most up to date data. There have been 2 new observations that are in the postnovkrat data (table 1) that should be added to the data.

In this clean data, the object ID represents the \# in columns it was in its original table, so it can be tracked back to the RAW data.

```{r}
#read excel spreadsheet edited from RAW data through excel
library(readxl)
clean_data_12_17 <-read_excel("cleanKratdata1217.xlsx")

```

Columns to be deleted: editor, creator, teeth

Columns to be edited:

Column: repro stage NA = non-repro

Column: life stage NA = adult

Column: recap NA = not recap AKA a new individual

Column: Tissue sample NA = no tissue sample was taken

column: Location = WILL NEED TO BE FIGURED OUT USING GPS COORDS – will either be "FORT ORD" or "HASTINGS" based on coordinates.

# Sorting out species by location

The two different locations that the data is from each have a different species. In Hastings Natural Reserve there is Dipodomys Venustus and in Fort Ord natural reserve there is Dipodomys Heermanni

Sorting species by the date I was collecting data at each location

11/8/25 HASTINGS

11/9/25 HASTINGS

11/22/25 HASTINGS

11/23/25 HASTINGS

7/13/25 FORT ORD

7/12/25 FORT ORD

6/22/25 HASTINGS

9/28/25 FORT ORD

10/18/25 FORT ORD

10/19/25 FORT ORD

```{r}
df <- clean_data_12_17
library(tidyverse)
#Create new data frame with a new column assigning the correct species by observation date
newdf <- df |> mutate(correct_species=case_when(
  CreationDate >= "2025-11-8"~"ve",
  CreationDate >= "2025-7-12" ~ "he",
  CreationDate <= "2025-6-22" ~ "ve"))
Checkingsort <- newdf |> select(correct_species, CreationDate)
Checkingsort
#CreationDate == c("2025-7-12","2025-7-13","2025-10-19"###,"2025-10-18","2025-9-28") ~ "he"))
#Create code to plot to compare between species and sex
#body length vs tail length
ggplot(data = newdf, mapping = aes(x=Body_Length, y= Tail_Length, color = correct_species))+geom_point()
#scale_color_manual(values = c("deeppink3","blue"))+
#geom_smooth()
  
  #labs(title = "Is there sexual dimorphism in bison?",x = "age (years)", y = "weight #(kg)",color = "Sex")
```

```{r}
#Foot length vs body length
ggplot(data = newdf, mapping = aes(x=Body_Length, y= Foot_Length, color = correct_species))+geom_point()
#max head width vs head length
ggplot(data = newdf, mapping = aes(x=Maxillary_Head_Width, y= Head_Length, color = correct_species))+geom_point()
#bolea head width vs head length
ggplot(data = newdf, mapping = aes(x=Bolea_Head_Width, y= Head_Length, color = correct_species))+geom_point()
#Hind limb length vs foot length
ggplot(data = newdf, mapping = aes(x=Hind_Limb_Length, y= Foot_Length, color = correct_species))+geom_point()
#finger length vs hand width
ggplot(data = newdf, mapping = aes(x=Finger_Length, y= Hand_Width, color = correct_species))+geom_point()
#finger length vs wrist width
ggplot(data = newdf, mapping = aes(x=Finger_Length, y= Wrist_Width, color = correct_species))+geom_point()
#Toe length vs foot width
ggplot(data = newdf, mapping = aes(x=Toe_Length, y= Foot_Width, color = correct_species))+geom_point()
#show foot width??? is the foot width missing a decimal in some places?!?!
newdf$Foot_Width
```

WE HAVE A PROBLEM!!! some of the foot width cells that have values in the 40s rather than 5.0 something are not foot width but instead they are forelimb length or foot length that has been misplaced. We will have to fix this in our dataset...

```{r}
#ear length vs head length
ggplot(data = newdf, mapping = aes(x=Ear_Length_Notch, y= Head_Length, color = correct_species))+geom_point()
```

```{r}
#seem to be catching much less males than females!!!!?!??
#make datasets that only include Males and then only include females due to krat sexual dimorphisms
newdf$Sex
maledf <- newdf |> filter(Sex == "Male")
femaledf <- newdf |> filter(Sex == "Female")
#male krats only ear length vs head length
ggplot(data = maledf , mapping = aes(x=Ear_Length_Notch, y= Head_Length, color = correct_species))+geom_point()

#female krats only ear-length vs head length
ggplot(data = femaledf , mapping = aes(x=Ear_Length_Notch, y= Head_Length, color = correct_species))+geom_point()
#female krats only--- wondering based on this graph if FORT ORD might have venustus???? tissue samples will be interesting!!! or else heermanni are just very variable esp.in these areas. There was a paper that said that FORT ORD would be good habitat for D. Venustus
#headlengthvs maxillary head width
  ggplot(data = femaledf , mapping = aes(x=Head_Length, y= Maxillary_Head_Width, color = correct_species))+geom_point()
#head length vs bolea head length
ggplot(data = femaledf , mapping = aes(x=Bolea_Head_Width, y= Head_Length, color = correct_species))+geom_point()
#bullae vs maxillary
ggplot(data = femaledf , mapping = aes(x=Bolea_Head_Width, y= Maxillary_Head_Width, color = correct_species))+geom_point()

#checking to make sure that species was sorted right into the correct_species column
newdf$correct_species
newdf$CreationDate
Checkingsort <- newdf |> select(correct_species, CreationDate)
Checkingsort
#figured out I missed a date at hastings wasnt sorted into species correctly--going back to fix it-- found typo where i said 7 and not 6 OOPS
```

11/8/25 HASTINGS

11/9/25 HASTINGS

11/22/25 HASTINGS

11/23/25 HASTINGS

7/13/25 FORT ORD

7/12/25 FORT ORD

6/22/25 HASTINGS

9/28/25 FORT ORD

10/18/25 FORT ORD

10/19/25 FORT ORD
