---
title: "Preliminary Models and Graphs"
author: "Mackenzie"
format: pdf
editor: visual
---

## Looking into the combined data frame
```{r}
library(tidyverse)
library(brms) 
library(ggeffects)
library(readxl)
```

## Make sure to run the code for the Data Frame final_krat_data


I don't think there is enough separation between species to be useful in this one
```{r}
# Foot Length on Toe Length
final_krat_data %>%
  ggplot(aes(x=Foot_Length, y = Toe_Length, color = Species))+
  geom_point()+geom_smooth(method = "lm")
```

This could be helpful but very few data points
```{r}
#Head Length on Bolea Head Width 
final_krat_data %>%
  ggplot(aes(x=Head_Length, y = Bolea_Head_Width, color = Species))+
  geom_point()+geom_smooth(method = "lm")
```


This shows a trend but has a very wide range, so I'm not sure that we can trust the association
```{r}
# Head Length on Maxillary Head Width
final_krat_data %>%
  ggplot(aes(x=Head_Length, y = Maxillary_Head_Width, color = Species))+
  geom_point()+geom_smooth(method = "lm")
```

Mostly very clustered points, doesn't show much
```{r}
# Head Length on Body Length
final_krat_data %>%
  ggplot(aes(x=Head_Length, y = Body_Length, color = Species))+
  geom_point()+geom_smooth(method = "lm")
```


Mostly very clustered points, doesn't show much
```{r}
#Head Length on Hind Limb Length 
final_krat_data %>%
  ggplot(aes(x=Head_Length, y = Hind_Limb_Length, color = Species))+
  geom_point()+geom_smooth(method = "lm")
```


I don't think this is very helpful 
```{r}
#Head length on Hand width
final_krat_data %>%
  ggplot(aes(x=Head_Length, y = Hand_Width, color = Species))+
  geom_point()+geom_smooth(method = "lm")
```

There aren't many venustus body lengths
```{r}
#Hind Limb Length on Body Length
final_krat_data %>%
  ggplot(aes(x=Hind_Limb_Length, y = Body_Length, color = Species))+
  geom_point()+geom_smooth(method = "lm")
```

I don't think there is enough data here
```{r}
#Foot Length on Toe Length
final_krat_data %>%
  ggplot(aes(x=Foot_Length, y = Toe_Length, color = Species))+
  geom_point()+geom_smooth(method = "lm")
```

This one doesn't seem helpful
```{r}
#Head length on Ear Notch
final_krat_data %>%
  ggplot(aes(x=Head_Length, y = Ear_Length_Notch, color = Species))+
  geom_point()+geom_smooth(method = "lm")+ylim(0,21)
```

I'm not sure what happened here, but I don't think we should use it
```{r}
#Wrist Width on Hand width
final_krat_data %>%
  ggplot(aes(x=Wrist_Width, y = Hand_Width, color = Species))+
  geom_point()+geom_smooth(method = "lm")
```


These next two could end up being useful. It seems like there are different trends between the species
```{r}
#Body length on Tail Length
final_krat_data %>%
  ggplot(aes(x=Body_Length, y = Tail_Length, color = Species))+
  geom_point()+geom_smooth(method = "lm")
```

```{r}
#Head length on Tail Length
final_krat_data %>%
  ggplot(aes(x= Head_Length, y = Tail_Length, color = Species))+
  geom_point()+geom_smooth(method = "lm")
```
```{r}
#Finger length vs Head length
final_krat_data %>%
  ggplot(aes(x=Head_Length, y = Finger_Length, color = Species))+
  geom_point()+geom_smooth(method = "lm")
```


# Testing Models
```{r}
#checking how R identified the columns
class(final_krat_data$Head_Length)
class(final_krat_data$Tail_Length)

#additive model for The effect of Head length by species on Tail length
m.head.tail.length.additive <-
brm(data = final_krat_data,
family = gaussian,
Tail_Length ~ 0 + Species +
Head_Length,
iter = 2000, warmup = 1000, chains = 4, cores = 4, seed = 4,
file = "output/m.head.tail.length.additive")

#Assessing the model
summary(m.head.tail.length.additive)
```
```{r}
plot(m.head.tail.length.additive)
```


```{r}
#Testing the interactive model of Head Length and Tail Length
m.head.tail.length.interactive <-
brm(data = final_krat_data,
family = gaussian,
bf(Tail_Length ~ 0 + a + b*Head_Length,
# Then, we specify that we want our intercept, a, to vary with species
a ~ 0 + Species,
# Next, we specify that we want our slope, b, to ALSO vary with species
# (this is the interaction part!!)
b ~ 0 + Species,
nl = TRUE),
iter = 2000, warmup = 1000, chains = 4, cores = 4,
seed = 4,
file = "output/m.head.tail.length.interactive")
```
```{r}
#assessing the interactive model of Head Length and Tail Length
summary(m.head.tail.length.interactive)
```
```{r}
plot(m.head.tail.length.interactive)
```
It looks like this model ran well, but the CI for both slopes includes zero
