---
title: "Testing predictive power"
format: pdf
editor: visual
---

## Goals

Compare real species with the species predicted by our measurements in the model.

Create a graph to show our real species values vs ones predicted by measurements

## Set up

```{r}
#load the packages necessary
library(ggplot2)
library(brms)
library(ggeffects)
library(tidyverse)
library(readxl)
library(dplyr)
#install.packages("ggtext")
library(ggtext)
library(here)
#Load the data
krat <- read.csv("final_krat_data_2026.csv")
#print(krat)

#Creating a new column to use as a proxy for the size of the individual by dividing head_length by body_length
df <- krat %>%
    mutate(relative_size = Body_Length / Head_Length)
```

### Testing the Finger length model's predictions:

```{r}

#Plot the Finger length vs Relative size
df %>%
  ggplot(aes(y= Finger_Length, x = relative_size, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")

#Running bernouilli model to predict species based on measurements
m.bern.finger <- brm(Species ~ 1 + Finger_Length + relative_size, 
                        data = df, 
                        family = bernoulli(), 
                        chains = 4, iter = 2000,
                        file = "output/m.bern.finger") 

#summary model
summary(m.bern.finger)
plot(m.bern.finger)

#Predict the response based on the model and plot it
preds.m.finger <- predict_response(m.bern.finger)
#plot this
plot(preds.m.finger, show_data = TRUE)
head(preds.m.finger$Finger_Length)
#Make model finger into its own data frame to plot
df.finger <- data.frame(preds.m.finger$Finger_Length)
head(df.finger)

#Create the plot from before but including error and prettier
pred.vs.finger <- df.finger |> ggplot(mapping=aes(y=predicted, x= x))+geom_point()+
#Add error bars using the confidence interval as the error bars
geom_errorbar(aes(ymin = conf.low, ymax = conf.high, color = std.error), width = 0.1)+
#Make it pretty
theme_bw()+
  theme(
    legend.text=element_text(face="italic"),
    plot.title=element_markdown(hjust=0.5))+
  scale_color_gradient(low= "black", high = "violetred")+labs(
    x="Finger length (mm)",
    y="Probability of Species",
    title="Species probability based on finger length values"
  )+
    labs(color = "Error")

ggsave("pred.vs.finger.png")

#Check the probability of each species
pp_check(m.bern.finger, type="scatter_avg")

#Use function to check if these were right
fitted(m.bern.finger)

```

How do I take the values from the preds.m.finger and create a new column in df?

```{r}
#GOAL: Create new column using probability estimates from preds.m.finger

#Create vector using values from model
Pred.sp <- fitted(m.bern.finger)
head(Pred.sp)
#Filter df for only observations including finger length AKA get rid of NAs
#df2 <- df |> filter(Finger_Length )
df_filtered <- df[!is.na(df$Finger_Length) & !is.na(df$relative_size), ]
nrow(df_filtered)
# Create as column
df_filtered$Probability.sp <- NA
head(df_filtered)

#Create new data frame using the vector
Pred.sp.data <- data.frame(Pred.sp)
head(Pred.sp.data$Estimate)
Pred.sp.est <- Pred.sp.data$Estimate

#showing rows of each
nrow(df_filtered)
length(Pred.sp.data$Estimate)
#Add data estimate to new column
df_filtered$Probability.sp <- Pred.sp.data$Estimate 
head(df_filtered)
#Should we add the error too???

#Here are the species predicted by the model versus the actual species shown in a plot for our presentation.
df_filtered |> ggplot(mapping = aes(y = Probability.sp, x = Species, color = Species))+geom_point()+geom_jitter(width = 0.2, height = 0)+theme_bw()+
  theme(
    legend.text=element_text(face="italic"),
    plot.title=element_markdown(hjust=0.5)
        )+
  scale_color_manual(values=c("darkturquoise","palevioletred"))+
  labs(
    x="Relative size (Head length (mm) / Body length (mm))",
    y="Finger length (mm)",
    title="Testing the correctness of species predictions based on finger length"
  )

#Violin plot option
df_filtered |> ggplot(mapping = aes(y = Probability.sp, x = Species, color = Species))+geom_violin()+geom_point()+geom_jitter(width = 0.2, height = 0)+theme_bw()+
  theme(
    legend.text=element_text(face="italic"),
    plot.title=element_markdown(hjust=0.5)
        )+
  scale_color_manual(values=c("darkturquoise","palevioletred"))+
  labs(
     x="Relative size (Head length (mm) / Body length (mm))",
    y="Finger length (mm)",
    title="Testing the correctness of species predictions based on finger length"
  )

#boxplot!!!!
pred.finger.boxplot <- df_filtered |> ggplot(mapping = aes(y = Probability.sp, x = Species, color = Species))+geom_boxplot()+geom_point()+geom_jitter(width = 0.2, height = 0)+theme_bw()+
  theme(
    legend.text=element_text(face="italic"),
    plot.title=element_markdown(hjust=0.5)
        )+
  scale_color_manual(values=c("darkturquoise","palevioletred"))+
  labs(
    x="Relative size (Head length (mm) / Body length (mm))",
    y="Finger length (mm)",
    title="Testing Accuracy of finger length based species predictions"
  )

ggsave("pred.finger.boxplot.png")
```

### Creating graphs to show how good finger measurement is at prediction

```{r}
#Goal: histogram of correct species vs species that would have been wrong if only based on model
#use casewhen to create species assumptions in model, =<50% prob is Dihe, >50% prob is Dive
df.pred.finger.m.sp <- df_filtered |> mutate(m.species = case_when(
  Probability.sp > 0.5 ~ "Dipodomys venustus",
  Probability.sp < 0.5 ~ "Dipodomys heermanni",
  Probability.sp == 0.5 ~ "unknown"
))
head(df.pred.finger.m.sp)


#plot the real species assignments
bar.real.sp.finger <- df.pred.finger.m.sp |> ggplot(mapping = aes(x=Species, fill = Species))+geom_bar()+theme_bw()+
  theme(
    legend.text=element_text(face="italic"),
    plot.title=element_markdown(hjust=0.5)
        )+
  scale_color_manual(values=c("darkturquoise","palevioletred"))+
  labs(
    x="Actual species",
    y="Count",
    title="Actual Species counts")+ labs(fill = "Species")

ggsave("bar.real.sp.finger.png")

#Plot the model species assignments
bar.pred.sp.finger <- df.pred.finger.m.sp |> ggplot(mapping = aes(x=m.species, fill = m.species))+geom_bar()+theme_bw()+
  theme(
    legend.text=element_text(face="italic"),
    plot.title=element_markdown(hjust=0.5)
        )+
  scale_color_manual(values=c("darkturquoise","palevioletred"))+
  labs(
    x="Species Prediction",
    y="Count",
    title="Species predictions based on finger length")+ labs(fill = "Predicted Species") 

ggsave("bar.pred.sp.finger.png")
```

## Testing hand width predictive capability

```{r}
#Plot the Hand width vs Relative size
df %>%
  ggplot(aes(y= Hand_Width, x = relative_size, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")

#Running bernouilli model to predict species based on measurements
m.bern.handW <- brm(Species ~ 1 + Hand_Width + relative_size, 
                        data = df, 
                        family = bernoulli(), 
                        chains = 4, iter = 2000,
                        file = "output/m.bern.handW") 

#summary model
summary(m.bern.handW)
plot(m.bern.handW)

#Predict the response based on the model and plot it
preds.m.handW <- predict_response(m.bern.handW)
plot(preds.m.handW, show_data = TRUE)

#Check the probability of each species
pp_check(m.bern.handW, type="scatter_avg")

#Use function to check if these were right
fitted(m.bern.handW)
```

```{r}
#GOAL: Create new column using probability estimates from preds.m.finger

#Create vector using values from model
Pred.sp.hand <- fitted(m.bern.handW)
head(Pred.sp.hand)
#Filter df for only observations including finger length AKA get rid of NAs
#df2 <- df |> filter(Finger_Length )
df.filt.hand <- df[!is.na(df$Hand_Width) & !is.na(df$relative_size), ]
nrow(df.filt.hand)
# Create as column
df.filt.hand$Probability.sp <- NA
head(df.filt.hand)

#Create new data frame using the vector
Pred.sp.hand.data <- data.frame(Pred.sp.hand)
head(Pred.sp.hand.data$Estimate)
Pred.sp.hand.est <- Pred.sp.hand.data$Estimate

#showing rows of each
nrow(df.filt.hand)
length(Pred.sp.hand.data$Estimate)
#Add data estimate to new column
df.filt.hand$Probability.sp <- Pred.sp.hand.data$Estimate 
head(df.filt.hand)
```

### Plots for Hand width species prediction

```{r}
#Here are the species predicted by the model versus the actual species shown in a plot for our presentation.
pred.handW.boxplot <- df.filt.hand |> ggplot(mapping = aes(y = Probability.sp, x = Species, color = Species))+geom_boxplot()+geom_point()+geom_jitter(width = 0.2, height = 0)+theme_bw()+ylim(0,1)+
  theme(
    legend.text=element_text(face="italic"),
    plot.title=element_markdown(hjust=0.5)
        )+
  scale_color_manual(values=c("darkturquoise","palevioletred"))+
  labs(
    x="Hand width size (mm)",
    y="Probability of species",
    title="Testing accuracy of hand width based species predictions"
  )

ggsave("pred.handW.boxplot.png")

#trying a violin plot
df.filt.hand |> ggplot(mapping = aes(y = Probability.sp, x = Species, color = Species))+geom_violin()+geom_point()+geom_jitter(width = 0.2, height = 0)+theme_bw()+ylim(0,1)+
  theme(
    legend.text=element_text(face="italic"),
    plot.title=element_markdown(hjust=0.5)
        )+
  scale_color_manual(values=c("darkturquoise","palevioletred"))+
  labs(
    x="Hand width size (mm)",
    y="Probability of species",
    title="Testing accuracy of hand width based species predictions"
  )
#trying strip plot
df.filt.hand |> ggplot(mapping = aes(y = Probability.sp, x = Species, color = Species))+geom_jitter(width = 0.2, height = 0)+theme_bw()+ylim(0,1)+
  theme(
    legend.text=element_text(face="italic"),
    plot.title=element_markdown(hjust=0.5)
        )+
  scale_color_manual(values=c("darkturquoise","palevioletred"))+
  labs(
    x="Hand width size (mm)",
    y="Probability of species",
    title="Testing accuracy of hand width based species predictions")
```

### Creating a graph which might better show how bad this measurement is

```{r}
#Goal: histogram of correct species vs species that would have been wrong if only based on model
#use casewhen to create species assumptions in model, <50% prob is Dihe, >50% prob is Dive
df.handW.m.pred <- df.filt.hand |> mutate(m.species = case_when(
  Probability.sp > 0.5 ~ "Dipodomys venustus",
  Probability.sp < 0.5 ~ "Dipodomys heermanni",
  Probability.sp == 0.5 ~ "unknown"
))
head(df.handW.m.pred)


#plot the real species assignments
bar.real.sp.handW <- df.handW.m.pred |> ggplot(mapping = aes(x=Species, fill = Species))+geom_bar()+theme_bw()+
  theme(
    legend.text=element_text(face="italic"),
    plot.title=element_markdown(hjust=0.5)
        )+
  scale_color_manual(values=c("darkturquoise","palevioletred"))+
  labs(
    x="Actual species",
    y="Count",
    title="Actual Species counts")+ labs(fill = "Species")

ggsave("bar.real.sp.handW.png")

#Plot the model species assignments
bar.pred.sp.handW <- df.handW.m.pred |> ggplot(mapping = aes(x=m.species, fill = m.species))+geom_bar()+theme_bw()+
  theme(
    legend.text=element_text(face="italic"),
    plot.title=element_markdown(hjust=0.5)
        )+
  scale_color_manual(values=c("darkturquoise","palevioletred"))+
  labs(
    x="Species Prediction",
    y="Count",
    title="Species predictions based on hand width")+ labs(fill = "Predicted Species") 

ggsave("bar.pred.sp.handW.png")
```
