---
title: "README: Preliminary Models and Graphs"
output: html_document
---

```{r message=FALSE, warning=FALSE}
#Necessary Packages and Data
library(tidyverse)
library(brms) 
library(ggeffects)
library(readxl)
library(here)

final_krat_data <- read.csv(here("output/final_krat_data_2026.csv"))
```

# Preliminary Models and Graphs
The Quarto documents in this folder show analysis done on the final_krat_data. 
(The cleaned data frame). Graphs were started by showing the relationship 
between two variables and separating by species. Some simple models were run on 
these graphs to quantify the relationships. An example of the code for a 
preliminary graph is shown below. Both Quarto documents are part of the 
exploratory analysis.



```{r warning=FALSE}
#Hind Limb Length on Body Length
final_krat_data %>%
  ggplot(aes(x=Hind_Limb_Length, y = Body_Length, color = Species))+
  geom_point()+geom_smooth(method = "lm")
```


## Initial Models
Some of the initial models used both the Gaussian and Categorical families in the brms package. 

```{reval=FALSE}
#Gaussian Family Model
m.head.tail.length.additive <-
brm(data = final_krat_data,
family = gaussian,
Tail_Length ~ 0 + Species +
Head_Length,
iter = 2000, warmup = 1000, chains = 4, cores = 4, seed = 4,
file = "output/m.head.tail.length.additive")

#Categorical Family Model
m.cat.head.tail <- brm(data = final_krat_data, 
                       formula = Species ~ Head_Length + Tail_Length,
                       family = categorical(link = "logit"), 
                       chains = 4, iter= 2000, 
                       file = "output/m.cat.head.tail")
```


## Updated Graphs
It was later determined to use Head Length divided by Body Length as a proxy for 
body size. The values given were compared to different variables and separated 
by species. An example graph code is shown below

```{r warning=FALSE}
#Head Length/Body_Length graph
final_krat_data %>%
  ggplot(aes(y=Head_Length/Body_Length, x = Body_Length, color =  Species))+
  geom_point()+
  geom_smooth(method = "lm")
```

## Updated Models
A new column called Relative_Head_Length was created for this proxy and was used 
to run models as the example below shows. The best morphological measurements 
for predictors of species were determined by these models.
```{reval=FALSE}
m.relhead.body <- brm(Species ~ Body_Length + Relative_Head_Length, 
                        data = divide_krat_data, 
                        family = bernoulli(), 
                        chains = 4, iter = 2000,
                        file = "output/m.relhead.body")
```

